---
title: "(Mortality)"
author: "Mustapha Ghazi"
date: "2023-02-25"
output: 
  html_document: 
    self_contained: FALSE
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(segmented)
library(seas)
library(readxl)
library(readr)
library(reshape2)
library(data.table)
library(mosaic)
library(modelr)
library(hexbin)
library(nortest)
library(rmarkdown)
library(dplyr)
library(kableExtra)
```

```{r}
setwd("~/Desktop/DATA")
#crudedata1_seer <- read.csv("crudedata1_seer.csv", header = T)
#sheet_1 <- read.csv("sheet 1.csv", header = T)
#sheet_3 <- read.csv("sheet 3.csv", header = T)

Leukemia_Mortality <- read.csv("Leukemia_Mortality.csv", header = TRUE)

#View(Leukemia_Mortality)
```

Some data cleaning

```{r, results = FALSE}
#base <- crudedata1_seer %>% filter(Sex != 0 & Year.of.death != 0) %>% dplyr::select(Age.recode.with..1.year.olds, Sex, Year.of.death)
#basicData <- base %>% rename("Age Recode (< 1)" = "Age.recode.with..1.year.olds", "Year of Death" = "Year.of.death"); basicData

leukemiaMortality <- Leukemia_Mortality %>% filter(State == 2 | State == 4 | State == 6 | State == 10 | State == 13 | State == 15 | State == 17 | State == 18 | State == 21 | State == 30 | State == 31 | State == 32 | State == 43 | State == 44 | State == 47); leukemiaMortality
```

Ages and their distributions

```{r}
ageCount <- leukemiaMortality %>% filter(!is.na(Count)) %>% count(Age.recode.with..1.year.olds) ; ageCount # ok so this isnt how many died but rather how many entries we have. Kinda useless unless you wanna descibe your dataset not the information conveyed

sum(ageCount$n) # see, matches the number of rows when leukemiaMortality %>% filter(!is.na(Count)) is done

deathAges <- leukemiaMortality %>% filter(!is.na(Count)) %>% dplyr::select(Age.recode.with..1.year.olds, Count) %>% group_by(Age.recode.with..1.year.olds) %>% count(Age.recode.with..1.year.olds, wt=Count) %>% rename("Age (recode with < 1 years old)" = "Age.recode.with..1.year.olds", "Count" = "n"); deathAges

n <- sum(deathAges$Count)

deathAges <- mutate(deathAges, Percent = 100*Count/n); deathAges

ggplot(deathAges, aes(deathAges$'Age (recode with < 1 years old)')) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Age (recode with < 1 years old)") + ggtitle("Deaths by Age Group")

ggplot(deathAges, aes(x = `Age (recode with < 1 years old)`, y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), hjust = -0.5)  + coord_flip() + 
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(labels = function(x) format(x, scientific=FALSE)) +
  ylab("Count") +
  xlab("Age (recode with < 1 years old)") +
  ggtitle("Deaths by Age Group")
```

Sex and their distributions

```{r}
sexCount <- leukemiaMortality %>% filter(!is.na(Count)) %>% count(Sex) ; sexCount # ok so this isnt how many died but rather how many entries we have. Kinda useless unless you wanna descibe your dataset not the information conveyed

sum(sexCount$n) # see, matches the number of rows when leukemiaMortality %>% filter(!is.na(Count)) is done

deathSex <- leukemiaMortality %>% filter(!is.na(Count)) %>% dplyr::select(Sex, Count) %>% group_by(Sex) %>% count(Sex, wt=Count) %>% rename("Count" = "n"); deathSex

n <- sum(deathSex$Count)

deathSex <- mutate(deathSex, Percent = 100*Count/n); deathSex

ggplot(deathSex, aes(Sex)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Sex") + ggtitle("Deaths by Sex Group")
```

State and their distributions

```{r}
stateCount <- leukemiaMortality %>% filter(!is.na(Count)) %>% count(State) ; stateCount # ok so this isnt how many died but rather how many entries we have. Kinda useless unless you wanna descibe your dataset not the information conveyed

sum(stateCount$n) # see, matches the number of rows when leukemiaMortality %>% filter(!is.na(Count)) is done

deathState <- leukemiaMortality %>% filter(!is.na(Count)) %>% dplyr::select(State, Count) %>% group_by(State) %>% count(State, wt=Count) %>% rename("Count" = "n"); deathState

n <- sum(deathState$Count)

deathState <- mutate(deathState, Percent = 100*Count/n); deathState

deathState$State <- as.character(deathState$State)

ggplot(deathState, aes(State)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("State") + ggtitle("Deaths by State") + coord_flip()
```

YOD and their distributions

```{r}
yodCount <- leukemiaMortality %>% filter(!is.na(Count)) %>% count(Year.of.death.recode) ; yodCount # ok so this isnt how many died but rather how many entries we have. Kinda useless unless you wanna descibe your dataset not the information conveyed

sum(yodCount$n) # see, matches the number of rows when leukemiaMortality %>% filter(!is.na(Count)) is done

deathYOD <- leukemiaMortality %>% filter(!is.na(Count)) %>% filter(Year.of.death.recode != 0) %>% dplyr::select(Year.of.death.recode, Count) %>% group_by(Year.of.death.recode) %>% count(Year.of.death.recode, wt=Count) %>% rename("Count" = "n"); deathYOD

n <- sum(deathYOD$Count)

deathYOD <- mutate(deathYOD, Percent = 100*Count/n); deathYOD

ggplot(deathYOD, aes(Year.of.death.recode)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Year of Death") + ggtitle("Deaths by Year of Death") + coord_flip()
```

Race and their distributions

```{r}
raceCount <- leukemiaMortality %>% filter(!is.na(Count)) %>% count(Race.recode..White..Black..Other.) ; raceCount # ok so this isnt how many died but rather how many entries we have. Kinda useless unless you wanna descibe your dataset not the information conveyed

sum(raceCount$n) # see, matches the number of rows when leukemiaMortality %>% filter(!is.na(Count)) is done

deathRace <- leukemiaMortality %>% filter(!is.na(Count)) %>% dplyr::select(Race.recode..White..Black..Other., Count) %>% group_by(Race.recode..White..Black..Other.) %>% count(Race.recode..White..Black..Other., wt=Count) %>% rename("Count" = "n"); deathRace

n <- sum(deathRace$Count)

deathRace <- mutate(deathRace, Percent = 100*Count/n); deathRace

ggplot(deathRace, aes(Race.recode..White..Black..Other.)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Race") + ggtitle("Deaths by Race") + coord_flip()
```

Exporting the new data #\`\`\`{r} leukemiaMortality2 \<-
read.csv("leukemiaMortality2.csv", header = TRUE); leukemiaMortality2
\<- leukemiaMortality2 %\>% mutate(Race = ifelse(Race == 0, "All",
ifelse(Race == 1, "White", ifelse(Race == 2, "Black", ifelse(Race == 3,
"Other", "Unspecified"))))) %\>% mutate(Age.Group = ifelse(Age.Group ==
0, "00 years", ifelse(Age.Group == 1, "01-04 years", ifelse(Age.Group ==
2, "05-09 years", ifelse(Age.Group == 3, "10-14 years", ifelse(Age.Group
== 4, "15-19 years", ifelse(Age.Group == 5, "20-24 years",
ifelse(Age.Group == 6, "25-29 years", ifelse(Age.Group == 7, "30-34
years", ifelse(Age.Group == 8, "35-39 years", ifelse(Age.Group == 9,
"40-44 years", ifelse(Age.Group == 10, "45-49 years", ifelse(Age.Group
== 11, "50-54 years", ifelse(Age.Group == 12, "55-59 years",\
ifelse(Age.Group == 13, "60-64 years", ifelse(Age.Group == 14, "65-69
years", ifelse(Age.Group == 15, "70-74 years", ifelse(Age.Group == 16,
"75-79 years", ifelse(Age.Group == 17, "80-84 years", ifelse(Age.Group
== 18, "85+ years", "Unknown")))))))))))))))))))) %\>%
mutate(Year.of.Death = ifelse(Year.of.Death == 0, "1969-2019",
ifelse(Year.of.Death == 1, "1969-1971", ifelse(Year.of.Death == 2,
"1972-1974", ifelse(Year.of.Death == 3, "1975-1977",
ifelse(Year.of.Death == 4, "1978-1980", ifelse(Year.of.Death == 5,
"1981-1983", ifelse(Year.of.Death == 6, "1984-1986",
ifelse(Year.of.Death == 7, "1987-1989", ifelse(Year.of.Death == 8,
"1990-1992", ifelse(Year.of.Death == 9, "1993-1995",
ifelse(Year.of.Death == 10, "1996-1998", ifelse(Year.of.Death == 11,
"1999-2001", ifelse(Year.of.Death == 12, "2002-2004",\
ifelse(Year.of.Death == 13, "2005-2007", ifelse(Year.of.Death == 14,
"2008-2010", ifelse(Year.of.Death == 15, "2011-2015",
"2016-2020"))))))))))))))))) %\>% mutate(Sex = ifelse(Sex == 0, "Male
and female", ifelse(Sex == 1, "Male", "Female"))) %\>% mutate(State =
ifelse(State == 2, "Arizona", ifelse(State == 4, "California",
ifelse(State == 6, "Connecticut", ifelse(State == 10, "Georgia",
ifelse(State == 13, "Illinois", ifelse(State == 15, "Iowa", ifelse(State
== 17, "Kentucky", ifelse(State == 18, "Louisiana", ifelse(State == 21,
"Massachusetts", ifelse(State == 30, "New Jersey", ifelse(State == 31,
"New Mexico", ifelse(State == 32, "New York", ifelse(State == 43,
"Texas",\
ifelse(State == 44, "Utah", "Washington")))))))))))))))

#leukemiaMortality2

#write.csv(leukemiaMortality2, "leukemiaMortalityDescription.csv",
row.names=FALSE) #\`\`\`

Redoing all the graphs with the Description data

```{r}
leukemiaMortalityDesc <- read.csv("leukemiaMortalityDescription.csv", header = TRUE);

deathAges2 <- leukemiaMortalityDesc %>% filter(!is.na(Count)) %>% filter(Race == "All") %>% filter(Sex == "Male and female") %>% dplyr::select(Age.Group, Count) %>% group_by(Age.Group) %>% count(Age.Group, wt=Count) %>% rename("Count" = "n"); deathAges2

n <- sum(deathAges2$Count)

deathAges2 <- mutate(deathAges2, Percent = 100*Count/n); deathAges2

ggplot(deathAges2, aes(x = Age.Group, y = Count, fill = Age.Group)) +
  geom_bar(stat = "identity") + coord_flip() +
  scale_y_continuous(labels = function(x) format(x, scientific=FALSE)) +
  ylab("Count") +
  xlab("Age (recode with < 1 years old)") + 
  ggtitle("Deaths by Age Group") + theme(legend.position="none")

ggplot(deathAges2, aes(x = Age.Group, y = Count/sum(Count), fill = Age.Group)) +
  geom_bar(stat = "identity") + coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  ylab("Density") +
  xlab("Age (recode with < 1 years old)") +
  theme(legend.position="none") +
  ggtitle("Deaths by Age Group")
```

```{r}
deathSex2 <- leukemiaMortalityDesc %>% filter(!is.na(Count)) %>% filter(Sex != 'Male and female') %>% filter(Race == "All") %>% dplyr::select(Sex, Count) %>% group_by(Sex) %>% count(Sex, wt=Count) %>% rename("Count" = "n"); deathSex2

n <- sum(deathSex2$Count)

deathSex2 <- mutate(deathSex2, Percent = 100*Count/n); deathSex2

ggplot(deathSex2, aes(Sex, fill = Sex)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Sex") + ggtitle("Deaths by Sex Group") + theme(legend.position="none")

ggplot(deathSex2, aes(Sex, fill = Sex)) + 
  geom_bar(aes(weight = Count/sum(Count))) +
  ylab("Density") + 
  xlab("Sex") + 
  ggtitle("Deaths by Sex Group") + 
  theme(legend.position="none")

```

```{r}
deathState2 <- leukemiaMortalityDesc %>% filter(!is.na(Count))  %>% filter(Race == "All") %>% filter(Sex == "Male and female") %>% dplyr::select(State, Count) %>% group_by(State) %>% count(State, wt=Count) %>%  rename("Count" = "n"); arrange(deathState2, Count)


n <- sum(deathState2$Count)

deathState2 <- mutate(deathState2, Percent = 100*Count/n); 
deathState2 <- arrange(deathState2, Count)

ggplot(deathState2, aes(reorder(State, (Count)), fill = State)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("State") + ggtitle("Deaths by State") + coord_flip() + theme(legend.position="none")

ggplot(deathState2, aes(reorder(State, (Count)), fill = State)) + 
  geom_bar(aes(weight = Count/sum(Count))) +
  ylab("Density") + 
  xlab("State") + 
  ggtitle("Deaths by State") + 
  coord_flip() + 
  theme(legend.position="none")

```

```{r}
deathYOD2 <- leukemiaMortalityDesc %>% filter(!is.na(Count))  %>% filter(Race == "All") %>% filter(Sex == "Male and female") %>% dplyr::select(Year.of.Death, Count) %>% filter(Year.of.Death != "1969-2019")  %>% group_by(Year.of.Death) %>% count(Year.of.Death, wt=Count) %>% rename("Count" = "n"); deathYOD2

n <- sum(deathYOD2$Count)

deathYOD2 <- mutate(deathYOD2, Percent = 100*Count/n); deathYOD2

ggplot(deathYOD2, aes(Year.of.Death, fill = Year.of.Death)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Year of Death") + ggtitle("Deaths by Year of Death") + coord_flip() + theme(legend.position="none")

ggplot(deathYOD2, aes(Year.of.Death, fill = Year.of.Death)) + 
  geom_bar(aes(weight = Count/sum(Count))) + 
  ylab("Density") + 
  xlab("Year of Death") + 
  ggtitle("Deaths by Year of Death") + 
  coord_flip() + 
  theme(legend.position="none")

```

```{r}
deathRace2 <- leukemiaMortalityDesc %>% filter(!is.na(Count)) %>% filter(Race != 'All')  %>% filter(Sex == "Male and female") %>% dplyr::select(Race, Count) %>% group_by(Race) %>% count(Race, wt=Count) %>% rename("Count" = "n"); arrange(deathRace2, Count)

n <- sum(deathRace2$Count)

deathRace2 <- mutate(deathRace2, Percent = 100*Count/n); arrange(deathRace2, Count)

deathRace2 %>%
  kbl(format="latex",
      col.names = c("Race","Count","Percent"),
      align="c") %>%
  kable_minimal(full_width = F,  html_font = "Source Sans Pro")

ggplot(deathRace2, aes(reorder(Race, (Count)), fill = Race)) + geom_bar(aes(weight = Count)) + ylab("Count") + xlab("Race") + ggtitle("Deaths by Race") + coord_flip() + theme(legend.position="none")

ggplot(deathRace2, aes(reorder(Race, (Count)), fill = Race)) + 
  geom_bar(aes(weight = Count/sum(Count))) +
  ylab("Density") + 
  xlab("Race") + 
  ggtitle("Deaths by Race") + 
  coord_flip() + 
  theme(legend.position="none")

```


